Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-07-05 trunk/papers/ArticleZ80/source_full/paper.tex
Ajustando o repositorio para artigo short

  This paper presents a formal model, developed in B, of the Z80
  instruction set. The formal model includes different libraries that
  specify entities commonly used in microcontrolers and
  microprocessors. Not only this model can be used for documentation
  purposes, but also to simulate and test the assembly code. It is also
  a fundamental component to extend the reach of the B method from the
  specification of the functional requirements of a system to an assembly-level
  implementation thereof.\newline %\keywords{Embedded Software, Simulation, B method and Verification}
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
 { \textbf{Key words:} B method, Simulation, Verification and Z80}

@@

\item[Power] contains the standard definition of exponentiation, as
  well as some basic lemmas, e.g. values of different powers of
  two, allproved interactively with the Atelier-B prover.
  This module is essential to establish the relationship between
  bit vectors and integer arithmetics. %%DD

@@

$$
\begin{array}{l}
\textit{bit\_not}  \in  \textit{BIT}  \fun  \textit{BIT}  \land 
\textit{bit\_not} \rm = \rm \{ \rm 0  \mapsto  \rm 1 , \rm 1  \mapsto  \rm 0 \rm \} \land \\
\forall (\textit{bb}).(\textit{bb} \in \textit{BIT} \implies \textit{bit\_not}(\textit{bit\_not}(\textit{bb})) = \textit{bb})
\end{array}
$$
Conjunction is a binary function on bits and it is defined as:
$$
\begin{array}{l}
\textit{bit\_and} \in \textit{BIT} \times \textit{BIT} \fun \textit{BIT} \land \\
\forall (\textit{b1}, \textit{b2}).(\textit{b1}  \in \textit{BIT}  \land \textit{b2} \in \textit{BIT} \implies \\
\quad ((\textit{bit\_and}(\textit{b1}, \textit{b2}) = 1) \iff (\textit{b1} = 1)  \land  (\textit{b2} = 1)))
\end{array}
$$
%The module provides the following lemmas for commutativity and associativity:
%$$
%\begin{array}{l}
%%  \textit{bit\_and}(0,0) = 0;  \textit{bit\_and}(0,1) = 0; \\
%%  \textit{bit\_and}(1,0) = 0;  \textit{bit\_and}(1,1) = 1; \\
%\forall (\textit{b1},\textit{b2}).(\textit{b1} \in \textit{BIT} \land \textit{b2} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{b2}) = \textit{bit\_and}(\textit{b2},\textit{b1})))\land \\
%\forall (\textit{b1},\textit{b2},\textit{b3}).(\textit{b1} \in \textit{BIT} \land  \textit{b2} \in \textit{BIT} \land \textit{b3} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{bit\_and}(\textit{b2},\textit{b3})) = \textit{bit\_and}(\textit{bit\_and}(\textit{b1},\textit{b2}),\textit{b3})))\\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 1) = \textit{b1})); \\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 0) = 0));
%\end{array}
%$$
The module provides lemmas for commutativity and associativity, definitions of $\textit{bit\_or}$ (disjunction)
and $\textit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their expression in B is
similar to $\textit{bit\_and}$.

@@

\hspace*{-0.0in}\it bv\_catenate  $\in$  \it BIT\_VECTOR  $\times$  \it BIT\_VECTOR  $\fun$ \it
BIT\_VECTOR $\land$\\
\hspace*{0.2in}\it bv\_catenate \rm =  $\lambda$  v\rm ,\it w \rm . \rm (\it v 
$\in$  \it BIT\_VECTOR $\land$ \it w $\in$  \it BIT\_VECTOR  $\mid$  \it v\^ \it w\rm ) %$\cat$
\newline

@@

Bytes, i.e. bit vectors of length 8, are a common entity in hardware
design. We provide the following definitions:

@@

\hspace*{0.0in}$\forall$ \rm (\it b1\rm )\rm .\rm (\it b1  $\in$  \it BYTE  $\implies$  \bf size\rm (\it b1\rm )\rm =\it BYTE\_WIDTH  $\land$  \it b1  $\in$  \bf seq1\rm (\it BIT\rm )\rm )

%\hspace*{0.0in}\it BYTE \rm = \rm \{ \it bt  $\mid$  \it bt $\in$ \it BIT\_VECTOR  $\land$  \it size\rm (\it bt\rm )\rm =\it BYTE\_WIDTH\rm \} $\land$

@@

Bit vectors are used to represent and combine numbers: integer ranges
(signed or unsigned). Therefore, our library includes functions to
manipulate such data, for example, the function $\textit{bv\_to\_nat}$
maps bit vectors to natural numbers:
$$
\begin{array}{l}
\textit{bv\_to\_nat} \in \textit{BIT\_VECTOR} \fun \nat  \land \\
\textit{bv\_to\_nat} = \lambda v .  (v \in \textit{BIT\_VECTOR} \mid  \sum i . (i \in \dom(v) . v(i)
\times 2^{i-1}))
\end{array}
$$

An important lemma associated to this definition is: 
$$\forall n . (n \in \nat_1 \implies \textit{bv\_to\_nat}(\textit{nat\_to\_bv}(n)) = n).$$

@@

Usually, each type module just needs to specialize definitions that
are given in the hardware modeling library.  For example, the function
$\textit{bv\_to\_nat}$ from bit vector arithmetic is specialized to
$\textit{byte\_uchar}$. As the set $\textit{BYTE}$ is a subset of the
$\textit{BIT\_VECTOR}$, this function can be defined as follows:
$$
\begin{array}{l}
\textit{byte\_uchar} \in \textit{BYTE} \fun \nat \land \\
\textit{byte\_uchar} = \lambda (v) . ( v \in BYTE | bv\_to\_nat(v) )
\end{array}
$$

@@

We also created lemmas on pairs of dual conversion functions, such as:
$$
 \begin{array}{l}
  \forall (val) . (val \in \textit{UCHAR} |
  \textit{byte\_uchar}(\textit{uchar\_byte}(val)) = val) \land\\
  \forall (by) . (by \in \textit{BYTE} |
  \textit{uchar\_byte}(\textit{byte\_uchar}(by)) = by)
 \end{array}
$$

@@

\item For the proof obligation (\it byte\_uchar\rm \ $\in$ \textit{BYTE} $\bij$
\textit{UCHAR}), the proof rule is: $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) \rm $\in$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )  $\bij$  \rm 0 $\upto$ \rm 2\rm 5\rm 5 \\

@@

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller
module Z80 and parts of it are presented below.  Basically, the
microcontroller state is formed by the program counter, ports,
registers, etc. The stae transition occurs either as an instruction
execution or an external action. Groups of registers are represented
by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables is in
the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined
by the clause \textit{OPERATIONS} and, in general, each instruction
has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
The functions specific to the microcontroller are defined in the ALU
(arithmetic logic unit) module. The main module includes an instance of
the memory module and accesses the definitions from basic data types
modules and the \textit{ALU} module.

@@

%%DD removi uma frase que não fazia sentido gramaticalmente.
The Z80 CPU contains a stack pointer ($\textit{sp}$), program counter
($\textit{pc}$), two index registers ($\textit{ix}$ and
$\textit{iy}$), an interrupt register ($\textit{i\_}$), a refresh
register ($\textit{r\_}$), two bits ($\textit{iff1}$, $\textit{iff2}$)
used to control the interruptions, a pair of bits to define the
interruption mode ($\textit{im}$) and the input and output ports
($\textit{i\_o\_ports}$). These definitions are represented by
\textit{INVARIANT}.

\begin{sloppypar}
\bf INVARIANT

@@

\hspace*{0.10in}\it iff2  $\in$  \it BIT  $\land$ \it im $\in$ (\it BIT $\times$ \it BIT\rm )  $\land$  \it i\_o\_ports  $\in$  \it BYTE  $\fun$  \it BYTE
\end{sloppypar}

% Features of microcontroller
The internal registers contains 176 bits of read/write memory that are
represented by identifiers used as parameters in the instructions. It
includes two sets of six general purpose registers which may be used
individually as 8-bit registers or as 16-bit register pairs.  The
working registers are represented by variable $\textit{rgs8}$. The
domain of $\textit{rgs8}$ ($\textit{id\_reg8}$) is a set formed by
identifiers of registers of 8 bits.  These registers are called main
register set (\textit{a0,b0,c0,d0,e0,f0,h0,l0}) and alternate register
set (\textit{a\_0,b\_0,c\_0,d\_0,e\_0,f\_0,h\_0,l\_0}).  These
registers can be accessed in pairs, forming 16-bits, resulting in
another set of identifiers of 16-bits registers, named
$\textit{id\_reg16}$. These sets are represented below.

@@

\bf SETS\\
\hspace*{0.10in}\it id\_reg\_8\rm =\rm \{\it a0\rm, \it f0\rm, \it b0\rm, \it
c0\rm, \it d0\rm, \it e0\rm, \it h0\rm, \it l0\rm, \it a\_0\rm,  \it f\_0\rm,
\it b\_0\rm, \it c\_0\rm, \it d\_0\rm, \it e\_0\rm, \it h\_0\rm, \it l\_0 \rm
\};\\ \hspace*{0.10in}\it id\_reg\_16\rm =\rm \{\it AF \rm, \it BC\rm, \it
DE\rm, \it HL\rm, \it SP\rm\}

@@

A related function is subtraction: $\textit{subtract8UCHAR}$. There
are also similar functions for the $\textit{SCHAR}$ type, they are
respectively $\textit{add8SCHAR}$ and $\textit{subtract8SCHAR}$, all
these functions are of 8 bits ($\textit{BYTE}$). In the same way, the
arithmetic functions for 16 bits ($\textit{BV16}$) are defined.  The
module ALU has several others functions, for
example:%but only some simplest functions also are explained below:
%\begin{itemize}
  %\item \it inc  $\in$ \it BYTE $\fun$  \it BYTE \rm - It receives a byte and
  %returns its increment. There is a similar function named $\textit{dec}$.
%  \item 
\it instruction\_next $\in$ USHORT $\fun$ USHORT \rm - It receives the
current value from program counter register (\textit{pc}) and returns
its increment;
% \item 
and \it is\_negative $\in$ \it BYTE $\fun$ \it BIT \rm - It returns
the most significant bit, in other words, the signal bit.
  % \item \it is\_zero  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns 1 if the received
 % byte is zero, otherwise returns 0.
  %\item \it update\_refresh\_reg \rm - It receives a byte and returns its increment until the seventh bit.
%\end{itemize}
As the logic functions that are defined in the \textit{BYTE} and
\textit{BV16} module are included in the \textit{ALU} module, they can
be seen and used directly in the \textit{ALU} and \textit{Z80}
modules.

@@

A simple example of instruction is $\textit{LD\_(nn)\_A}$\footnote{The
  B tools does not allow o use parentheses in identifiers, and the
  characters \{``('',``)''\} are replaced respectively by
  \{``9'',``0''\} in the actual specification.} shown in this
section. The pre-defined functions are necessary many times to model
the instructions, these functions facilitate the construction of
instruction set model. By default, all parameters in operations are
either predefined elements in the model or integers values in decimal
representation. This instruction use the $\textit{updateAddressMem}$
operation from \textit{Memory} module, it receives an address memory
and its new memory value. $\textit{LD\_(nn)\_A}$ also increments the
program counter ($\textit{pc}$) and updates the refresh register
($\textit{r\_}$). The other instructions have a similar structure.

\hspace*{0.00in}\bf LD\_9nn0\_A \rm ( \it nn \rm ) \rm =

@@

\hspace*{0.00in}\bf END\rm

>>>>

  This paper presents a formal model, developed in B, of the Z80
  instruction set. The formal model includes different libraries that
  specify entities commonly used in microcontrolers and
  microprocessors. Not only this model can be used for documentation
  purposes, but also to simulate and test the assembly code. It is also
  a fundamental component to extend the reach of the B method from the
  specification of the functional requirements of a system to an assembly-level
  implementation thereof.
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
% { \textbf{Key words:} B method, Simulation, Verification and Z80}

@@

\item[Power] contains the standard definition of exponentiation, as
  well as some basic lemmas, e.g. values of different powers of
  two, all proved interactively with the Atelier-B prover.
  This module is essential to establish the relationship between
  bit vectors and integer arithmetics. %%DD

@@

$$
\hspace*{-1.2in}\begin{array}{l}
\textit{bit\_not}  \in  \textit{BIT}  \fun  \textit{BIT}  \land 
\textit{bit\_not} \rm = \rm \{ \rm 0  \mapsto  \rm 1 , \rm 1  \mapsto  \rm 0 \rm \} \land \\
\forall (\textit{bb}).(\textit{bb} \in \textit{BIT} \implies \textit{bit\_not}(\textit{bit\_not}(\textit{bb})) = \textit{bb})
\end{array}
$$
Conjunction is a binary function on bits and it is defined as:
$$
\hspace*{-1.4in}\begin{array}{l}
\textit{bit\_and} \in \textit{BIT} \times \textit{BIT} \fun \textit{BIT} \land \\
\forall (\textit{b1}, \textit{b2}).(\textit{b1}  \in \textit{BIT}  \land \textit{b2} \in \textit{BIT} \implies \\
\quad ((\textit{bit\_and}(\textit{b1}, \textit{b2}) = 1) \iff (\textit{b1} = 1)  \land  (\textit{b2} = 1)))
\end{array}
$$
%The module provides the following lemmas for commutativity and associativity:
%$$
%\begin{array}{l}
%%  \textit{bit\_and}(0,0) = 0;  \textit{bit\_and}(0,1) = 0; \\
%%  \textit{bit\_and}(1,0) = 0;  \textit{bit\_and}(1,1) = 1; \\
%\forall (\textit{b1},\textit{b2}).(\textit{b1} \in \textit{BIT} \land \textit{b2} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{b2}) = \textit{bit\_and}(\textit{b2},\textit{b1})))\land \\
%\forall (\textit{b1},\textit{b2},\textit{b3}).(\textit{b1} \in \textit{BIT} \land  \textit{b2} \in \textit{BIT} \land \textit{b3} \in \textit{BIT} \implies \\
%\quad (\textit{bit\_and}(\textit{b1}, \textit{bit\_and}(\textit{b2},\textit{b3})) = \textit{bit\_and}(\textit{bit\_and}(\textit{b1},\textit{b2}),\textit{b3})))\\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 1) = \textit{b1})); \\
%% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 0) = 0));
%\end{array}
%$$
The module provides lemmas for commutativity and associativity, definitions of $\textit{bit\_or}$ (disjunction)
and $\textit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their representations in B are
similar to $\textit{bit\_and}$.

@@

\hspace*{-0.0in}\it bv\_catenate  $\in$  \it BIT\_VECTOR  $\times$  \it BIT\_VECTOR  $\fun$ \it
BIT\_VECTOR $\land$\\
\hspace*{0.2in}\it bv\_catenate \rm =  $\lambda$  v\rm ,\it w \rm . \rm (\it v 
$\in$  \it BIT\_VECTOR $\land$ \it w $\in$  \it BIT\_VECTOR  $\mid$  \it v\^ \it w\rm ) %$\cat$
\vspace*{0.05in}

@@

Bytes, i.e. bit vectors of length 8, are a common entity in hardware
design. We provide the following definitions:\\

@@

\hspace*{0.0in}$\forall$ \rm (\it b1\rm )\rm .\rm (\it b1  $\in$  \it BYTE  $\implies$  \bf size\rm (\it b1\rm )\rm =\it BYTE\_WIDTH  $\land$  \it b1  $\in$  \bf seq1\rm (\it BIT\rm )\rm ) 
\\
%\hspace*{0.0in}\it BYTE \rm = \rm \{ \it bt  $\mid$  \it bt $\in$ \it BIT\_VECTOR  $\land$  \it size\rm (\it bt\rm )\rm =\it BYTE\_WIDTH\rm \} $\land$

@@

Bit vectors are used to represent and combine numbers: integer ranges
(signed or unsigned). Therefore, our library includes functions to
manipulate such data, for example, the function $\textit{bv\_to\_nat}$
maps bit vectors to natural numbers:
$$
\hspace*{-0.3in}\begin{array}{l}
\textit{bv\_to\_nat} \in \textit{BIT\_VECTOR} \fun \nat  \land \\
\textit{bv\_to\_nat} = \lambda v .  (v \in \textit{BIT\_VECTOR} \mid  \sum i . (i \in \dom(v) . v(i)
\times 2^{i-1}))
\end{array}
$$

An important lemma associated to this definition is: 
$$\hspace*{-1.59in}\forall n . (n \in \nat_1 \implies \textit{bv\_to\_nat}(\textit{nat\_to\_bv}(n)) = n).$$

@@

Usually, each type module just needs to specialize definitions that
are given in the hardware modeling library.  For example, the function
$\textit{bv\_to\_nat}$ from bit vector arithmetic is specialized to
$\textit{byte\_uchar}$. As the set $\textit{BYTE}$ is a subset of the
$\textit{BIT\_VECTOR}$, this function can be defined as follows:
$$
\begin{array}{l}
\textit{byte\_uchar} \in \textit{BYTE} \fun \nat \land 
\textit{byte\_uchar} = \lambda (v) . ( v \in BYTE | bv\_to\_nat(v) )
\end{array}
$$

@@

We also created lemmas on pairs of dual conversion functions, such as:
$$
\hspace*{-0.7in} \begin{array}{l}
  \forall (val) . (val \in \textit{UCHAR} |
  \textit{byte\_uchar}(\textit{uchar\_byte}(val)) = val) \land\\
  \forall (by) . (by \in \textit{BYTE} |
  \textit{uchar\_byte}(\textit{byte\_uchar}(by)) = by)
 \end{array}
$$

@@

\item For the proof obligation (\it byte\_uchar\rm \ $\in$ \textit{BYTE} $\bij$
\textit{UCHAR}), the proof rule is: $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) \rm $\in$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )  $\bij$  \rm (0 $\upto$ \rm 2\rm 5\rm 5) \\

@@

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller
module Z80 and parts of it are presented below.  Basically, the
microcontroller state is formed by the program counter, ports,
registers, etc. The state transition occurs either as an instruction
execution or an external action. Groups of registers are represented
by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables is in
the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined
by the clause \textit{OPERATIONS} and, in general, each instruction
has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
The functions specific to the microcontroller are defined in the ALU
(arithmetic logic unit) module. The main module includes an instance of
the memory module and accesses the definitions from basic data types
modules and the \textit{ALU} module.

@@

%%DD removi uma frase que não fazia sentido gramaticalmente.
The Z80 CPU contains a stack pointer ($\textit{sp}$), program counter
($\textit{pc}$), two index registers ($\textit{ix}$ and
$\textit{iy}$), an interrupt register ($\textit{i\_}$), a refresh
register ($\textit{r\_}$), two bits ($\textit{iff1}$, $\textit{iff2}$)
used to control the interruptions, a pair of bits to define the
interruption mode ($\textit{im}$) and the input and output ports
($\textit{i\_o\_ports}$). These definitions are represented by
\textit{INVARIANT}.\\

\begin{sloppypar}
%\bf INVARIANT

@@

\hspace*{0.10in}\it iff2  $\in$  \it BIT  $\land$ \it im $\in$ (\it BIT $\times$ \it BIT\rm )  $\land$  \it i\_o\_ports  $\in$  \it BYTE  $\fun$  \it BYTE\\
\end{sloppypar}

% Features of microcontroller
The internal registers contains 176 bits of read/write memory that are
represented by identifiers used as parameters in the instructions. It
includes two sets of six general purpose registers which may be used
individually as 8-bit registers or as 16-bit register pairs.  The
working registers are represented by variable $\textit{rgs8}$. The
domain of $\textit{rgs8}$ ($\textit{id\_reg8}$) is a set formed by
identifiers of registers of 8 bits.  These registers are called main
register set (\textit{a0,b0,c0,d0,e0,f0,h0,l0}) and alternate register
set (\textit{a\_0,b\_0,c\_0,d\_0,e\_0,f\_0,h\_0,l\_0}).  These
registers can be accessed in pairs, forming 16-bits, resulting in
another set of identifiers of 16-bits registers, named
$\textit{id\_reg16}$. These sets are represented by  \textit{SETS}.\\

@@

%\bf SETS\\
\hspace*{0.0in}\it id\_reg\_8\rm =\rm \{\it a0\rm,\it f0\rm,\it b0\rm,\it c0\rm,\it d0\rm,\it e0\rm,\it h0\rm,\it l0\rm,\it a\_0\rm,  \it f\_0\rm,\it b\_0\rm,\it c\_0\rm,\it d\_0\rm,\it e\_0\rm,\it h\_0\rm,\it l\_0 \rm\};\\ 
\hspace*{0.2in}\it id\_reg\_16\rm =\rm \{\it AF\rm,\it BC\rm,\it
DE\rm,\it HL\rm,\it SP\rm\}\\

@@

%There is a related function $\textit{subtract8UCHAR}$ and others
A related function is subtraction: $\textit{subtract8UCHAR}$. There
are also similar functions for the $\textit{SCHAR}$ type, for example,
$\textit{add8SCHAR}$ and  $\textit{subtract8SCHAR}$. All
these functions are of 8 bits ($\textit{BYTE}$). In the same way, the
arithmetic functions for 16 bits ($\textit{BV16}$) are defined.  The
module ALU has several others functions, for
example: %but only some simplest functions also are explained below:
%\begin{itemize}
  %\item \it inc  $\in$ \it BYTE $\fun$  \it BYTE \rm - It receives a byte and
  %returns its increment. There is a similar function named $\textit{dec}$.
%  \item 
\it instruction\_next $\in$ USHORT $\fun$ USHORT \rm - It receives the
current value from program counter register (\textit{pc}) and returns
its increment;
% \item 
and \it is\_negative $\in$ \it BYTE $\fun$ \it BIT \rm - It returns
the most significant bit, in other words, the signal bit.
  % \item \it is\_zero  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns 1 if the received
 % byte is zero, otherwise returns 0.
  %\item \it update\_refresh\_reg \rm - It receives a byte and returns its increment until the seventh bit.
%\end{itemize}
As the logic functions that are defined in the \textit{BYTE} and
\textit{BV16} module are included in the \textit{ALU} module, they can
be seen and used directly in the \textit{ALU} and \textit{Z80}
modules.

@@

A simple example of instruction is $\textit{LD\_(nn)\_A}$\footnote{The
  B tools does not allow o use parentheses in identifiers, and the
  characters \{``('',``)''\} are replaced respectively by
  \{``9'',``0''\} in the actual specification.} shown in this
section. The pre-defined functions are necessary many times to model
the instructions, these functions facilitate the construction of
instruction set model. By default, all parameters in operations are
either predefined elements in the model or integers values in decimal
representation. This instruction use the $\textit{updateAddressMem}$
operation from \textit{Memory} module, it receives an address memory
and its new memory value. $\textit{LD\_(nn)\_A}$ also increments the
program counter ($\textit{pc}$) and updates the refresh register
($\textit{r\_}$). The other instructions have a similar structure.
\\ \\
\hspace*{0.20in}\bf LD\_9nn0\_A \rm ( \it nn \rm ) \rm =

@@

\hspace*{0.00in}\bf END\rm 
\\

%\subsection{Modeling the input/output instructions}
%\label{sec:modelingIO}
%%The Z80 has an extensive set of input and output (I/O) instructions and 256 ports for
%%devices. This model can transfer data blocks between the I/O devices and any
%%of the internal registers or memory address. The $\textit{IN\_r(C)}$ instruction is
%%represented by the following B operation. It receives a register identifier ``$\textit{rr}$''
%%and, it stores the value from ``rr'' in the $\textit{C}$ port address. Besides,
%%it increments the program counter and updates the flag registers.
%%
%%\hspace*{0.0in}\bf IN\_r\_9C0 \rm ( \it rr \rm ) \rm =
%%
%%\hspace*{0.0in}\bf PRE \it rr  $\in$  \it id\_reg\_8  $\land$  \it rr  $\not =$  \it f0\hspace*{0.15in}\bf THEN
%%
%%\hspace*{0.20in}\bf ANY \hspace*{0.10in}\it negative \rm , \it zero \rm , \it half\_carry \rm , \it pv \rm , \it add\_sub \rm , \it carry
%%
%%\hspace*{0.20in}\bf WHERE 
%%
%%\hspace*{0.40in}\it negative $\in$ \it BIT $\land$ \it zero $\in$ \it BIT $\land$ \it half\_carry $\in$ \it BIT 
%%$\land$ \it pv $\in$ \it BIT $\land$
%%
%% \hspace*{0.40in}\it add\_sub $\in$ \it BIT $\land$ \it carry $\in$ \it BIT  $\land$
%%
%%\hspace*{0.40in}\it negative \rm = \it is\_negative \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$ 
%%
%%\hspace*{0.40in}\it zero \rm = \it is\_zero \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$ \it half\_carry \rm = \rm 0  $\land$ 
%%
%%\hspace*{0.40in}\it pv \rm = \it parity\_even \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) $\land$
%%\it add\_sub \rm =\hspace*{0.10in}\rm 0 % $\land$ \it carry \rm = \it z\_c
%%
%%\hspace*{0.20in}\bf THEN
%%
%%\hspace*{0.40in}\it rgs8 \rm := \it rgs8  $\lover$  \rm \{ \rm ( \it rr  $\mapsto$  \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) \rm ,
%%
%%\hspace*{0.40in}\it update\_flag\_reg\rm (\it negative\rm,\it zero\rm,\it half\_carry\rm,\it pv\rm,\it add\_sub\rm,\it carry)\rm\}$\para$
%%
%%\hspace*{0.40in}\it pc \rm := \it instruction\_next \rm ( \it pc \rm )  $\para$  \it r\_ \rm := \it update\_refresh\_reg\rm (\it r\_\rm )
%%
%%\hspace*{0.20in}\bf END
%%
%%\hspace*{0.0in}\bf END \rm
%%
%%%The main input and output instructions work similarly, for example: $\textit{OUT(n),A}$  or $\textit{OUT (C ), r}$. 
%%
%% $
%% \begin{array}{l} 
%% \textit{IN\_r\_9C0} ( rr ) = \\
%% \quad   \PRE rr \in id\_reg\_8    \THEN\\
%% \quad\quad \ANY data\_in, negative , zero , half\_carry , pv , add\_sub , carry\\ 
%% \quad\quad \WHERE data\_in \in \textit{BYTE} \land negative \in \textit{BIT}\land\\
%% \quad\quad\quad carry \in \textit{BIT} \land half\_carry \in \textit{BIT} \land zero \in \textit{BIT} \land \\
%% \quad\quad\quad negative = is\_negative (data\_in) \land zero = is\_zero(data\_in ) \land\\
%% \quad\quad\quad half\_carry = 0 \land pv =parity\_even\_BYTE ( data\_in )    \land\\
%% \quad\quad\quad add\_sub =  0 \land carry = z\_c \\
%% \quad\quad  \THEN \\
%% \quad\quad\quad i\_o\_ports ( rgs8 ( c0 ) ) := data\_in ||\\
%% \quad\quad\quad rgs8 := rgs8 <+ \{ ( rr |-> data\_in ) ,\\
%% \quad\quad\quad get\_new\_flag\_register\_SZ\_H\_PvNC ( rgs8 , negative , zero, half\_carry , pv , add\_sub , carry ) \} ||\\
%% \quad\quad\quad pc := instruction\_next( pc )\\
%% \quad\quad \END\\
%% \quad \END\\
%% \end{array}
%% $
%
%
%
%
%%\subsection{Modeling the external actions}
%%\label{sec:externalactions}
%
The external actions change the state of the microcontroller, for
example, refreshing the I/O ports and requesting interruptions. The
external actions are also modeled by operations and they are named
with the prefix ``$ext\_$'' and followed by the name of action. There
are four external actions: $ext\_update\_io\_ports$, $ext\_NMI$ and
$ext\_INT$, $ext\_Reset$. The instruction $ext\_update\_io\_ports$
just changes the state of I/O port, see.\\
